# **Logistic regression**

__Summary of week3 study__


+ _In week3, I self-studied some data wrangling skills (join,mutate etc.), logistic regression, cross validation from  [DataCamp](https://campus.datacamp.com/courses/helsinki-open-data-science/logistic-regression?ex=1);_

+ _Logistic regression is about modeling categorical targets using related variables_

```{r}
## Loading packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally)
```

## 1. Read the data
In this week, we will use logisitic regression to build a model to predict alcohol consumption (AC). The data is processed from the [online data](https://archive.ics.uci.edu/ml/datasets/Student+Performance). There are two questionaires in the website. I have used 13 variables of both (`school`,`sex`,`age`,`address`,`famsize`,`Pstatus`,`Medu`,`Fedu`,`Mjob`,`Fjob`,`reason`,`nursery`,`internet`) to join the data. The remaining data is combined as mean if they are numeric or kept if they are other data types.

```{r}
## Reading data to alc 
data <- "/Users/qingli/Documents/GitHub/IODS-project/Data/processed_alc_data_w3.csv"
alc <- read.csv(data)
colnames(alc)
str(alc)
```

The explanations for the variables:

+ `school` -  student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)
+`sex` - student's sex (binary: 'F' - female or 'M' - male) 
+ `age` - student's age (numeric: from 15 to 22) 
+ `address` - student's home address type (binary: 'U' - urban or 'R' - rural) 
+ `famsize` - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3) 
+ `Pstatus` - parent's cohabitation status (binary: 'T' - living together or 'A' - apart) + `Medu` - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 = 5th to 9th grade, 3 = secondary education or 4 = higher education) 
+ `Fedu` - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2  =  5th to 9th grade, 3 =  secondary education or 4 = higher education) 
+ `Mjob` - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other') 
+ `Fjob` - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other') 
+ `reason` - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other') 
+ `guardian` - student's guardian (nominal: 'mother', 'father' or 'other') 
+ `traveltime` - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour) 
+ `studytime` - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours) 
+ `failures` - number of past class failures (numeric: n if 1<=n<3, else 4) 
+ `schoolsup` - extra educational support (binary: yes or no) 
+ `famsup` - family educational support (binary: yes or no) 
+ `paid` - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no) 
+ `activities` - extra-curricular activities (binary: yes or no) 
+ `nursery` - attended nursery school (binary: yes or no) 
+ `higher` - wants to take higher education (binary: yes or no) 
+ `internet` - Internet access at home (binary: yes or no) 
+ `romantic` - with a romantic relationship (binary: yes or no) 
+ `famrel` - quality of family relationships (numeric: from 1 - very bad to 5 - excellent) 
+ `freetime` - free time after school (numeric: from 1 - very low to 5 - very high) 
+ `goout` - going out with friends (numeric: from 1 - very low to 5 - very high) 
+ `Dalc` - workday alcohol consumption (numeric: from 1 - very low to 5 - very high) 
+ `Walc` - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high) 
+ `health` - current health status (numeric: from 1 - very bad to 5 - very good) 
+ `absences` - number of school absences (numeric: from 0 to 93)

These grades are related with the course subject, Math or Portuguese: 

+ `G1` - first period grade (numeric: from 0 to 20) 
+ `G2` - second period grade (numeric: from 0 to 20) 
+ `G3` - final grade (numeric: from 0 to 20, output target)

## 2. Select alcohol consumption (AC) related variables

To predict the alcohol consumption, I think the folowing variables might be associated:

+ `paid`: the extra money can help to buy alcohol.
+ `goout`: there will be more chance to drink alcohol with friends if they go out very often.
+ `famrel`: people with higher AC are unlikely to have bad family relationships.
+ `absences`: number of school absences is a sign for who drinks often and cannot make to school.

## 3. Explore the relationship between the selected variables with alcohol consumption (AC)

```{r}
## high_use vs paid
alc %>% group_by(paid,high_use) %>% summarise(count=n())
```

The student got extra paid or not seemed not affect AC.

```{r,fig.height=3,fig.width=4}
## high_use vs goout
high_AC_goout_mean <- mean(alc$goout[alc$high_use == "TRUE"])
low_AC_goout_mean <- mean(alc$goout[alc$high_use == "FALSE"])

print(c("high AC group goout_mean",high_AC_goout_mean,
        "low AC group goout_mean",low_AC_goout_mean))

g1 <- ggplot(alc, aes(x = high_use, y = goout,col=paid)) + geom_boxplot() +
  ggtitle("goout by high_use") + ylab("go_out freq") +xlab("achohol consumption")

g1
```

The students who have higher AC are more frequently going out with friends. The mean goout frequency for higher AC group is 3.7 compared with 2.8 in lower AC group.

```{r,fig.height=3,fig.width=4}
## high_use vs family relationship
high_AC_famrel_mean <- mean(alc$famrel[alc$high_use == "TRUE"])
low_AC_famrel_mean <- mean(alc$famrel[alc$high_use == "FALSE"])
print(c("high AC group family relationship status:",
        high_AC_famrel_mean,"low AC group family relationship status:",low_AC_famrel_mean))
g2 <- ggplot(alc, aes(x = high_use, y = famrel)) + geom_boxplot() +
 ggtitle("family relationship by high_use") +ylab("family relationship status") +xlab("achohol consumption")
g2
```

The family relationship status in students with higher AC and lower AC does not show a big difference from the mean of both group (3.7 and 4.0, respectively). But it looks different from the boxplot.

```{r,fig.height=3,fig.width=4}
## high_use vs absence
high_AC_absence_mean <- mean(alc$health[alc$high_use == "TRUE"])
low_AC_absence_mean <- mean(alc$health[alc$high_use == "FALSE"])

print(c("high_AC group absence times (mean):",high_AC_absence_mean,
        "low_AC group absence times (mean):",low_AC_absence_mean))
g3 <- ggplot(alc, aes(x = high_use, y = absences, col=paid)) +
 geom_boxplot() + ggtitle("absence by high_use") +ylab("absence times") +xlab("achohol consumption")
g3
```

The absence times in students with higher AC and lower AC does not show a big difference from the mean of both group (3.7 and 3.5, respectively). It slightly changed in boxplot, where the higher AC group shows more absence times.

## 4. Logistic regression

```{r}
m <- glm(high_use ~ paid + goout + famrel + absences, data = alc, family = "binomial")
summary(m)
```
+ `Intercept`: -2.56121, is the log odds of a student with higher AC when other variables are zero/paid.no.
+ The logistic regression model shows `goout` and `absences` are the most two most important explanatory variables in the model with significant low `p-value`. `famrel` follows them with less significant `p-value`. Surprisingly, `paid.yes` not does contribute to AC that much.
+  Coefficients for the variables (e.g. 0.775 for `goout`) is the expected change in log odds for a one-unit increase of `goout` in this case.
+  coefficient for `paiedyes` is measured by `Wald test` which takes `yes` in categorical variable `paid` as the reference class compared to those are not. 


```{r}
odd_ratio <- coef(m) %>% exp
CI <- confint(m) %>% exp
cbind(odd_ratio,CI)
```
+ The odds ratio for `Intercept`(0.07) means the odds of being higher AC is 0.07 times of the odds being lower AC when other variables are zero.
+ The odds ratio for `paidyes` means the odds for `paidyes` are about 1.35 times of the odds for `paidno`, and in the range of values [0.83100065 2.2091855] that you can be 95% certain to find the true mean of the OR.
+ The odds ratios and 95 confidence intervals are listed for other three numeric variables (e.g. `goout`). The odds of being higher AC is 2.17 times higher than the odds of being lower AC if we increase goout a one-unit higher. The range of [1.72550406 2.7679697] is that we can be 95% certain to find the true mean of the OR.

Therefore, we are only considering the significantly contirbuted variables in our new model which will be used for prediction.
```{r}
m_new <- glm(high_use ~ goout + famrel + absences, data = alc, family = "binomial")
summary(m_new)
```

## 5. Model prediction
```{r,fig.height=3,fig.width=5}
## use the new model to predict the probabilities for each sample
probabilities <- predict(m_new, type = "response")

## add probabilities and prediction to alc
alc <- mutate(alc, probabilities = probabilities)
alc <- mutate(alc, prediction = probabilities>0.5)

## calculate 2x2 cross tabulation of predictions versus the actual values 
table(high_use = alc$high_use, prediction = alc$prediction)

## calculate 2x2 cross tabulation of predictions versus the actual values in proportion
ggplot(alc, aes(x = probabilities, y =high_use,col = prediction)) + geom_point()
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table %>% addmargins
```
From the prediction result, there are 75.6% (`0.63874346+0.11780105=0.7565445`) of the samples out of `382` samples are accuretely predicted by our model.

```{r}
## the mean prediction error
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5 ## the class is either 'TRUE'(1) or 'FALSE' (0). 
  mean(n_wrong)
}
## call the loss function to calculate the wrong prediction rate
loss_func(class = alc$high_use, prob = alc$probabilities)
```
Thus, the error prediction rate is around 24.3%, which matches with our accuracy rate `1-accuracy rate(75.6%)`.